Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004]. 
This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
More recent work has included a focus not only on light, but also on leaf height in the canopy, as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007]. 
Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004; @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020]. 
As canopy shade is known to alter these trait relationships [@osnasDivergentDriversLeaf2018], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.

>>>>

Anatomical, structural and biochemical leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, metabolism, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004]. 
Vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
More recent work has extended the focus from light alone to leaf height in the canopy, as important in determining differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [Table `r table_leaf_traits`, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
Overall, leaves may optimize their biochemistry, anatomy and structure for their local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007]. 
Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004; @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020]. 
As canopy shade is known to alter these trait relationships [@osnasDivergentDriversLeaf2018], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
